Servants or Syndicates?: The Rise of Domestic Political Interventionism by Senior Military Leaders (Prospectus)

### 2020-11-18 code/prospectusv2.Rmd
commit

---
title: "Saviors or Syndicates?: The Rise of Domestic Political Interventionism by Retired Military Leaders (Prospectus)"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |

  _This dissertation explores the changes in how a powerful block of military professionals - retired senior military leaders - intervene in domestic contemporary politics. I advance a theory which argues that changes in political activism by retired senior leaders is a result of two interacting factors: the level of domestic political polarization within a society, and the degree of prestige and credibility that military leaders enjoy within the society. To support my argument, I trace changes in levels and types of distinct forms of political activism, including authoring opinion-editorials in major newspapers, and participation in presidential endorsements. I further advance the theory by conducting a survey experiment that tests for differences between civilian and military responses after exposure to situations that evoke one's partisan or ideological identity. I ultimately argue that the increasing political activism of retired military leaders is not merely a response to the hyper-partisanship present in the current political atmosphere, but rather the logical outworking of trends that have been active for some time in which both the ideological differences between Americans and the prestige of the military have increased. I conclude by describing the effect of the implications of growing politicla activism by senior military leaders on the core beliefs of the military regarding its very own essence and role in society._

@@

The central argument made in this dissertation is that increasing political activism at the hands of retired military officers is the result of two longstanding and interconnected trends: rising political polarization within the domestic political sphere, and rising prestige and credibility of military institutions afforded by American society. The interaction of these trends explains the remarkable increase in political activity by retired senior military leaders, and portends a shift in the perceived role that these senior military leaders fill vis-a-vis civilian authority - a shift away from professional and obedient servants and towards "guardians of democracy" who pick and choose when and how to intervene in American politics. Such a shift builds on previous work regarding normative "role beliefs" [@fitch_armed_1998] of the military institution, and has significant ramifications on the organizational culture of the military, its relationship with government and society, and American national security.

@@

Civil-military relations involve a wide host of significant and important theoretical questions that involve the pull and push of power between institutions, the delineation of authority between military and civilian leaders, and a number of other consequential issues. At its root, the study of civil-military relations examines what Peter Feaver calls the "problematique," or how to ensure that the military is strong enough to defeat threats to the polity without threatening the polity itself [@feaver_civil-military_1996; @feaver_civil-military_1999]. Feaver describes this "problematique" as "vexing" and a great "paradox" in that the military must be "strong enough to prevail in war," yet submissive to authority such that it cannot and will not "destroy or prey on the society it is intended to protect" [@feaver_civil-military_1999, p. 214]. Within the context of the US, the study of civil-military relations often focuses on the relationship between military and government elites, or what Risa Brooks calls "leadership at the state's apex" [@brooks_risa_integrating_2019, p. 380]. Indeed, it is at this level that vital decisions are made, including to go to war, how to prosecute war, how to achieve and implement effective strategy, how to prepare for the next conflict, and how large the military and its budget should be. A number of scholars have offered solutions to Feaver's "problematique," but none has been more lasting and consequential than the late Samuel Huntington, whose 1957 seminal work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, often serves as the point of departure for many scholarly works on civil-military relations. Although there have been many critiques of Huntington over the years [for example, see @janowitz_professional_1960; @finer_man_1962; @cohen_supreme_2003; @feaver_armed_2003; @brooks_paradoxes_2020], it is absolutely essential to understand Huntington's concept of normative civil-military relations, and what lies at the heart of it - a theory of civilian control of the military that maximizes the concept of what Huntington called "military professionalism" [@huntington_soldier_1957].

@@

Numerous critiques surrounding the Huntingtonian conception of _military professionalism_ have been levied over the years. Most of these critiques, in one way or another, charge that a strict theoretical and practical separation between military and political operating spheres, which Huntington to a certain degree certainly argued for, is fundamentally flawed. Many, if not all, of these critiques cite Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605], and thus point out that by its very nature, war requires the highest integration - not segregation - of political and military activities. Moreover, these critiques often point out, correctly in this author's opinion, that the conduct of war is more complicated than civilians simply telling the military the goals it wishes to achieve, and then allowing the military to determine how best to accomplish these goals. Rather, the development and implementation of strategy, especially during a time of war, requires an "iterative" dialogue between military leaders and civilians to ensure that effective policies are adopted [@rapp_civil-military_2015, 20].

Others argue that Huntington's conception of military professionalism is too narrow to be of practical use in the contemporary period in that such a model actually enables, rather than constrains, inappropriate political activity by military officers [@brooks_paradoxes_2020]. Risa Brooks charges that military officers who strongly identify with an ethic that stresses remaining apolitical at all costs, which she alleges follows from a Huntingtonian conception of military professionalism, may develop "blind spots such that they fail to recognize the political content or impact of their actions" [@brooks_paradoxes_2020, 17]. Whether military leaders recognize it or not, Brooks argues, military leaders comprise a profession that is inherently political, and such a reality ought to be acknowledged and embraced [@brooks_paradoxes_2020]. Regardless of whether and the degree to which one agrees or disagrees with Huntington's concept of military professionalism, a significant number of scholars note that a significant issue involves attempting to define where the line between appropriate political expression and inappropriate partisan activity should be drawn [@owens_military_2015; @golby_beyond_2015; @brooks_paradoxes_2020]. It is a difficult task for many reasons because, as Mackubin Thomas Owens points out, civil-military relations in the United States can be viewed as a "bargain" [@owens_us_2011] between the American people, the military, and the government that has to account for constantly-changing "political, social, technological, or geopolitical changes" [@owens_is_2016]. In other words, the ground is always moving for the actors involved in civil-military relations.

## The Dependent Variable: Military Intervention in Politics and Military "Role Beliefs"

Given the normative model espoused by Huntington and the critiques his model has been subjected to, it is important to understand what frameworks exist for understanding why the military would intervene in politics. Numerous scholars, mainly writing from a comparative perspective (although not exclusively), have enriched our understanding of what possibly motivates the military to interject itself into the political sphere. To be clear, these scholars are focusing mainly on active-duty military organizations, and not on the specific block of retired military officers, as I plan to examine, but the motivations and mechanisms for intervention that they put forward enrich our understanding of the phenomenon of political intervention by the military.

The British political scientist, S.E. Finer, argues that the military's intervention in politics is a function of the opportunity and desire to do so [@finer_man_1962]. Finer identifies several factors which should normatively inhibit the military from intervening in politics. These factors are military professionalism, the principle of civilian supremacy, and fear that the military would fail to remain a cohesive fighting force should it fracture internally upon intervening in political matters [@finer_man_1962]. Finer critiques the concept of Huntingtonian military professionalism somewhat harshly, arguing that although military professionalism should act to focus the military on itself and thus not on political activities, those who claim to be professional might and have used the concept of professionalism as an excuse to not obey civilian authorities [@finer_man_1962]. Finer is clear that the principle of civilian control - not military professionalism - is the true and only bulwark that can keep the military out of politics.

Describing the various motives that militaries might have for intervening in politics, Finer goes on to describe at least three. He first lists the motive of the "manifest destiny of the soldiers," in which the military acts as "saviors" of the country [@finer_man_1962, 28-29]. This motive, Finer claims, arises because of the confluence of the military's unique mission, sense of power, and claim of being "outside party politics" [@finer_man_1962, 29]. A second motive that militaries might possess which disposes them to intervene in politics is what Finer calls the "unique identification with the national interest" [@finer_man_1962, 30]. Acting from such a motive, Finer conceptualizes the military acting as "custodians" of the "national interest" [@finer_man_1962, 30]. Finer acknowledges that "custodianship" can take on different forms in different contexts, but points out that some militaries might view "custodianship" as a "duty to arbitrate or veto...to exercise it when some convulsion or decision of the civil autorities seems to them [the military] to threaten what they think are the permanent interests of the nation" [@finer_man_1962, 31]. The final motive Finer spends the most time on, calling it "the sectional interest," which refers to a more narrow subset of interests than the "national interest," and includes a wide ranging set of interests based on social class, regional affiliation, personal career motivations, or organizational interests [@finer_man_1962, 34-53].

Ultimately, Finer develops a typology of four levels, or degress of, military intervention. The first he calls "influence," and argues that at this level of intervention, the military makes appeals "to reason or emotion" in an attempt to persuade the civilian government [@finer_man_1962, 77]. Finer adds that intervening at this level is, in his mind at least, not problematic and entirely "consistent with the supremacy of the civil power" [@finer_man_1962, 77]. The next level of intervention is "pressures or blackmail," describing efforts by the military to "convince the civil power by threat of some sanction" [@finer_man_1962, 77]. Finer adds that military intervention via this level is "on the margin," referring to his belief that while some cases may be unproblematic, many other forms, such as the "threat to stage a coup," would be normatively wrong [@finer_man_1962, 77]. Finer points out a third level of intervention, that of "displacement," referring to the "removal" of government officials by the military as a result of "violence or the threat of violence" [@finer_man_1962, 78]. The fourth level, which is basically displacement on a wider scale, Finer calls "supplantment," which "sweeps away the civilian regime and establishes the military in its place" [@finer_man_1962, 78]. Finer goes on to explain the level of intervention as a function of the level of "political culture" that is prevalent in a state and society  [do I need more here? See Finer, pp. 79 and following]

Another helpful conceptualization of military intervention in politics is the idea of "role beliefs," a concept articulated by the political scientist, J. Samuel Fitch, in his book, _The Armed Forces and Democracy in Latin America._

A third, and equally rich, study of the degrees of military intervention in politics is available from Brian Taylor's book, _Politics and the Russian Army: Civil-Military Relations, 1689-2000.

@@

The question of if and how political polarization drives changes in the professional identity of the military spans several related literatures, most notable of which are American Politics and International Relations. Broadly speaking, this project is focused on how one aspect of domestic politics, political polarization, impacts a specific but important aspect of international relations, civil-military relations, which itself spans a wide and diverse literature. Because I intend to focus my work on the military as an institution, my dissertation will also rely on literature from the fields of sociology, organizational psychology, and ethics. I will frame my research question within each of these fields briefly, and to define important concepts as required, before addressing why this question matters and should be studied in depth.

Within the American Politics literature, there is broad recognition that political polarization levels in contemporary America are higher now than at any point since the end of the American Civil War [@hare_polarization_2014], but there are multiple explanations for what has caused this trend, ranging from a growing divide over matters pertaining to race and religion [@abramowitz_great_2018], the size and role of government [@sides_two_2015], and disagreements over specific policy issues [@mann_its_2012]. Despite the wide-ranging scholarship exploring the many facets of political polarization, virtually all scholars agree that contemporary levels of political polarization in the United States are historically abnormal and pose myriad problems.

One of the major concerns regarding political polarization is its on impact on American foreign policy in a broad sense, and specifically, on America's ability to lead and impact the world. Kenneth Schultz, for instance, underscored four challenges that have emerged as a result of America's polarized domestic politics, including the ability to obtain bipartisan support for "risky undertakings," the ability to learn the "lessons of foreign policy failure," the inability to "make long term commitments to allies and adversaries" as a result of "dramatic policy swings from one administration to another," and the "vulnerability" of America's electoral system to foreign third party adversaries [@schultz_perils_2017]. Because the military is a central actor in shaping America's foreign policy, obtaining a better understanding of how political polarization shapes various military institutions will yield findings that may help policymakers and academics help solve the problems Schultz identifies.

# Where My Puzzle Fits and Why It Matters

@@

One potential area of work that remains under-developed, both theoretically and empirically, is an exploration of the relationship between polarizing domestic politics and the professional identity and character of the military. I am specifically interested in the question of if and how political polarization impacts the concept of loyalty in the military. This is an interesting research question because it brings together at least three relevant strands of literatures which, when combined, generates an interesting set of puzzles and questions. First, one of the most fundamental normative hallmarks of the military's professional nature, at least by Huntingtonian standards, is that it ought to be an apolitical and non-partisan institution. Second, one of the features of the US military is that it prizes and elevates the trait of loyalty among those who serve and, by statute as well as normative tradition, exercises simultaneous dual loyalties to its civilian heads, the President as the Commander in Chief, and Congress. And third, we know from the American politics literature that ideological polarization and partisan loyalty are connected in that those citizens who hold to stronger ideological convictions tend to be the most loyal partisans in terms of voting and other political behaviors [@abramowitz_great_2018, 112-116].

@@

The first way in which the apolitical and non-partisan ethos of the military is challenged as a result of increasing political polarization is that there is increasing difficulty for individuals to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of military professionalism. This happens because as "conflict extension" [@layman_party_2002] occurs, an increasing number of issues are brought into partisan battlegrounds, and thus, there is simply less politically-neutral territory available. Thus, an actor such as the military, who may wish to remain non-partisan and out of politics to the extent that it can, may simply be less able to do so.

A second challenge military leaders and the institution faces in a polarized climate is from the rise of "negative partisanship"[@abramowitz_great_2018] that dominates the polarized American political landscape. As citizens hold increasingly strong and negative feelings towards those who side differently on a host of issues, there may be a tendency for members of the military institution to harbor similar feelings and attitudes. Even if the impact of this feature on the military and its leaders, due to training and professional development that emphasizes the importance of upholding apolitical and non-partisan behavioral norms is somewhat muted, the point is that members of the military institution are likely experiencing the same phenomenon - even if it is to a lesser degree - as the country's citizens. One important question about "negative partisanship" in a military setting, though, is whether service members hold strong negative feelings toward only civilians who are on the opposite side of the ideological spectrum, or whether such feelings also permeate within the ranks.

>>>>

---
title: "Servants or Syndicates?: The Rise of Domestic Political Interventionism by Senior Military Leaders (Prospectus)"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |

  _This dissertation explores changes in how the contemporary military institution intervenes in domestic politics. I advance a theory which argues that normative beliefs of military officers regarding political activism is a result of two interacting factors: the level of domestic political polarization within a society, and the degree of prestige and credibility the military holds. To support my argument, I trace changes in levels and types of distinct forms of political activism by military officers, including authoring opinion-editorials in major newspapers, as domestic polarization and military institutional credibility rises. I ultimately argue that the increasing political activism of retired military leaders is not merely a response to the hyper-partisanship present in the current political atmosphere, but rather the logical outworking of historical trends that have been building for decades. I conclude by describing the implications of changing beliefs among military leaders regarding the role the military should have in domestic politics. This dissertation contributes to the fields of international relations and civil-military relations by advanving an empirical and historical analysis of a growing phenomenon: military leaders intervening in domestic American politics. _

@@

The central argument made in this dissertation is that increasing political activism at the hands of military officers is the result of two longstanding and interconnected trends: rising political polarization within the domestic political sphere, and rising prestige and credibility of military institutions afforded by American society. The interaction of these trends explains the remarkable increase in political activity by retired senior military leaders, and portends a shift in the perceived role that these senior military leaders fill vis-a-vis civilian authority - a shift away from professional and obedient servants and towards "guardians of democracy" who pick and choose when and how to intervene in American politics. Such a shift builds on previous work regarding normative "role beliefs" [@fitch_armed_1998] of the military institution, and has significant ramifications on the organizational culture of the military, its relationship with government and society, and American national security.

@@

My proposed dissertation topic features within the subfield of civil-military relations, a broad academic subfield that is generally placed within the International Relations and Comparative Politics subfield (though within the American context, the subfield of American Politics also certainly applies). Civil-military relations involve a wide host of significant and important theoretical questions that involve the pull and push of power between institutions, the delineation of authority between military and civilian leaders, and the relationship between a state's military and society.

At its root, the study of civil-military relations examines what Peter Feaver calls the "problematique," or the question of how to ensure that a military is strong enough to defeat threats to the polity without threatening the polity itself [@feaver_civil-military_1996; @feaver_civil-military_1999]. Feaver describes this "problematique" as "vexing" and a great "paradox" in that the military must be "strong enough to prevail in war," yet submissive to authority such that it cannot and will not "destroy or prey on the society it is intended to protect" [@feaver_civil-military_1999, p. 214]. Moreover, the study of civil-military relations often focuses on the relationship between military and government elites, or at what Risa Brooks calls "leadership at the state's apex" [@brooks_risa_integrating_2019, p. 380]. Indeed, it is at this level of interaction between government and military leaders that vital decisions are made, including decisions whether to go to war, how to achieve and implement effective strategy, how to prepare for the next conflict, and how large the military and its budget should be.

A number of scholars have offered solutions to the"problematique" raised by Feaver, but none has been more lasting and consequential than the late Samuel Huntington, whose 1957 seminal work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, addresses this fundamental problem. Indeed, even more than 60 years after he wrote his book, Huntington's work often serves as the point of departure for many scholarly works on topics concerning civil-military relations. Although there have been many critiques of Huntington over the years [for example, see @janowitz_professional_1960; @finer_man_1962; @cohen_supreme_2003; @feaver_armed_2003; @brooks_paradoxes_2020], it is absolutely essential to understand Huntington's normative conception of effective civil-military relations, and what lies at the heart of it - a theory of civilian control of the military that maximizes a concept that Huntington called "military professionalism" [@huntington_soldier_1957].

@@

Numerous critiques surrounding the Huntingtonian conception of _military professionalism_ have been levied over the years. Most of these critiques, in one way or another, charge that a strict theoretical and practical separation between military and political operating spheres, which Huntington to a certain degree certainly argued for, is fundamentally flawed. Many, if not all, of these critiques cite Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605], and thus point out that by its very nature, war requires the highest integration - not segregation - of political and military activities. Moreover, these critiques often point out, correctly in this author's opinion, that the conduct of war is more complicated than civilians simply telling the military the goals it wishes to achieve, and then allowing the military to determine how best to accomplish these goals. Rather, the development and implementation of strategy, especially during a time of war, requires an "iterative" dialogue between military leaders and civilians to ensure that effective policies are adopted [@rapp_civil-military_2015, 20]. Others argue that Huntington's conception of military professionalism is too narrow to be of practical use in the contemporary period in that such a model actually enables, rather than constrains, inappropriate political activity by military officers [@brooks_paradoxes_2020]. Risa Brooks charges that military officers who strongly identify with an ethic that stresses remaining apolitical at all costs, which she alleges follows from a Huntingtonian conception of military professionalism, may develop "blind spots such that they fail to recognize the political content or impact of their actions" [@brooks_paradoxes_2020, 17]. Whether military leaders recognize it or not, Brooks argues, military leaders comprise a profession that is inherently political, and such a reality ought to be acknowledged and embraced [@brooks_paradoxes_2020].

Regardless of whether and the degree to which one agrees or disagrees with Huntington's concept of military professionalism, a significant number of scholars note that an enduring issue involves attempting to define where the line between appropriate political expression and inappropriate partisan activity should be drawn [@owens_military_2015; @golby_beyond_2015; @brooks_paradoxes_2020]. It is a difficult task for many reasons because, as Mackubin Thomas Owens points out, civil-military relations in the United States can be viewed as a "bargain" [@owens_us_2011] between the American people, the military, and the government that has to account for constantly-changing "political, social, technological, or geopolitical changes" [@owens_is_2016]. In other words, the ground is always moving for the actors involved in civil-military relations.

# Framing the DV: Military Intervention in Politics and Military "Role Beliefs"

Given Huntington's normative model and the critiques his model has been subjected to, it is important to understand what frameworks exist for understanding the debates surrounding why the military would or should intervene in politics, and to what degree. Here again, the literature is vast. Scholars from the field of comparative politics have generally tackled the question of why militaries intervene in political matters, likely as a result of the variation that is available in country-specific or cross-national studies. It is vital to briefly understand a couple of these important frameworks.

The British political scientist, S.E. Finer, argues that military intervention in politics is ultimately a function of both the opportunity and desire to do so [@finer_man_1962]. Finer first identifies several factors that should inhibit the military from intervening in politics, including the concept of military professionalism (as espoused by Huntington), strict adherence to the principle of civilian supremacy, and the military's fear that intervention in politics would lead to it fracturing internally, and thus failing to remain a cohesive fighting force [@finer_man_1962]. Finer delivers a rather stinging critique of Huntingtonian military professionalism, and argues that although professionalism should work to focus the military on itself and thus, away from politics, the danger of the concept of military professionalism is that it can also be abused, particularly by military leaders who adopt the view that they serve the larger concept of "the state" rather than a particular civilian government [@finer_man_1962, 22-23]. Finer argues that military leaders throughout history, including von Seeckt and Gröner under the German Weimar Republic, and MacArthur under Truman in the United States, inappropriately intervened in politics precisely because their conceptions of "professionalism" demanded that they did so [@finer_man_1962], a critique of military professionalism that Risa Brooks shares [@brooks_paradoxes_2020]. Finer's whole point is that the principle of civilian control - not military professionalism - is and ought to serve as the only bulwark to keep the military out of politics.

Finer further describes at least three motives that militaries might have for intervening in politics.  He argues that soldiers might adopt a specific form of "manifest destiny," or a worldview in which the military acts as "saviors" of the country [@finer_man_1962, 28-29]. This motive, Finer claims, arises because of the confluence of the military's unique mission, sense of power, and claim of being "outside party politics" [@finer_man_1962, 29]. A second motive that militaries might possess which disposes them to intervene in politics is what Finer calls the "unique identification with the national interest" [@finer_man_1962, 30]. Finer argues that militaries which act from such a motive might view themselves as the "custodians" of the "national interest" [@finer_man_1962, 30], a role that can take different forms, including a "duty to arbitrate or veto...to exercise it when some convulsion or decision of the civil authorities seems to them [the military] to threaten what they think are the permanent interests of the nation" [@finer_man_1962, 31]. Finally, Finer posits that militaries intervene in politics because of "the sectional interest," a wide-ranging set of reasons based on social class, regional affiliation, personal career motivations, or organizational interests [@finer_man_1962, 34-53].

Finer further classifies military intervention as one of four ascending levels. At the bottom or first level, which he calls "influence," Finer views the military as making appeals "to reason or emotion" in an attempt to persuade the civilian government to adopt a certain course of action [@finer_man_1962, 77]. Finer adds that intervention by the military at this level is not problematic and entirely "consistent with the supremacy of the civil power" [@finer_man_1962, 77]. The next level of intervention Finer labels "pressures or blackmail," describing efforts by the military to "convince the civil power by threat of some sanction" [@finer_man_1962, 77]. Finer adds that military intervention via this level is "on the margin," referring to his belief that while some cases may be unproblematic, many other forms, such as the "threat to stage a coup," are normatively wrong [@finer_man_1962, 77]. The third level of intervention, "displacement," refers to the "removal" of government officials by the military as a result of "violence or the threat of violence," and the fourth level, "supplantment," occurs when the military "sweeps away the civilian regime and establishes the military in its place" [@finer_man_1962, 78].

Another helpful conceptualization of military intervention in politics is the concept of military "role beliefs," introduced by J. Samuel Fitch in his book, _The Armed Forces and Democracy in Latin America._ Through rich interview data conducted in the 1990s, Fitch traces how leaders within the militaries of Argentina and Ecuador adopted one of five distinct "role beliefs," which describe the perceptions held by these officers of their normative role in politics [@fitch_armed_1998, 61]. The first type of role belief is "democratic professionalism," and involves both "a subjective commitment to democracy as the preferred political system," and a "belief in a professional ethic that differentiates and separates military and political affairs" [@fitch_armed_1998, 66]. A second "role belief" is "classical professionalism," and differs from the first in that while it "argues for the subordination of the military, in both political and policy terms, to the constitutionally designated civilian authorities", it does not espouse a "normative commitment to a democratic regime" [@fitch_armed_1998, 66-67]. A third type of "role belief" that military offices can adopt is that of a "power factor," or a belief that within the regime, "the military should have some degree of influence in deference to its vital societal role" [@fitch_armed_1998, 67]. The fourth type of "role belief" militaries can adopt is one of "conditional separation" to civilian leaders, and refers to the notion that while the military should stay out of politics "in normal circumstances", the advent of "crisis situations" can and does warrant the military to "safeguard the highest interests of the nation," and, as Fitch adds, "even if that entails overthrowing the incumbent government and assuming power temporarily" [@fitch_armed_1998, 68]. A final possible role belief militaries can adopt is a "tutelary" model in which the civilian government grants the military both a "high degree of autonomy" and the ability to become a regular and "institutional voice in policy matters" [@fitch_armed_1998, 68-69]. According to this "role belief," the military believes itself to be the final "guardian of national security," and thus that it has "the right" to participate in a wide variety of policy debates that impact the state [@fitch_armed_1998, 69]. Fitch argues that the particular "role beliefs" militaries adopt are a function of three primary factors: the lessons militaries draw from previous domestic intervention in politics, a wide-ranging assessment of current domestic indicators (the economy, the effectiveness of institutions, etc.), and the prevailing  international context [@fitch_armed_1998].

Within the US context, scholars have long been interested in the specific political activities that the military engages in. These studies have tended to examine specific political behaviors such as voting [@dempsey_our_2010], social media use [@urben_like_2017], political endorsements [@griffiths_not_2019], and attitudes towards political engagement [@urben_party_2013; @urben_wearing_2014], but they have not attempted to systematically explore these changes over time. Additionally, while Risa Brooks has created a helpful typology of political activities the US military engages in, ranging from "public appeals" that engage the public, to "shoulder tapping" focused exclusively on Congress [@brooks_militaries_2009], her typology also does not explain variation in these activities over time. While the domestic environment in Latin America is vastly different than that in the United States, Fitch's work underscores and points to the importance of civilian and military leaders committing themselves to the supremacy of both democratic institutions and democratic rule [@fitch_armed_1998]. And while there has arguably been more variation in Latin American militaries intervening in politics, I believe that there is sufficient evidence that indicates a subtle but important shift in the US military's beliefs regarding its central role and purpose has occurred, and that a systematic examination of this shift is warranted.

# Framing the IVs: Political Polarization and Military Prestige

@@

Within the American Politics literature, there is broad recognition that political polarization levels in contemporary America are higher now than at any point since the end of the American Civil War [@hare_polarization_2014], and there are multiple explanations for what has caused this trend, ranging from a growing divide over matters pertaining to race and religion [@abramowitz_great_2018], the size and role of government [@sides_two_2015], and disagreements over specific policy issues [@mann_its_2012]. Despite the wide-ranging scholarship within the field of American Politics that explores the facets of political polarization, virtually all scholars agree that contemporary levels of political polarization pose myriad problems.

Scholars of international relations have, more recently, examined the pernicious impacts of polarization on foreign policy. Kenneth Schultz, for instance, underscored four challenges that have emerged as a result of America's polarized domestic politics, including the ability to obtain bipartisan support for "risky undertakings," the ability to learn the "lessons of foreign policy failure," the inability to "make long term commitments to allies and adversaries" as a result of "dramatic policy swings from one administration to another," and the "vulnerability" of America's electoral system to foreign third party adversaries [@schultz_perils_2017]. Because the military is a central actor in shaping America's foreign policy, obtaining a better understanding of how political polarization shapes various military institutions will yield findings that may help policymakers and academics help solve the problems Schultz identifies.

But a more central concern of mine is understanding how and why political polarization might impact how the military views its central role in American politics. I argue that there are at least three ways through which increasing polarization can shape the apolitical and non-partisan ethos of the military, and thus, can alter how the military views its role in politics. The first is that there is increasing difficulty for military leaders to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of Huntingtonian military professionalism. Fewer and fewer issue areas and policy matters are outside of the "spell" of polarization, owing to a phenomenon schoalrs of American politis refer to as "conflict extension" [@layman_party_2002]. In previous eras of polarization, whereas a single issue may have served as the "battleground" for partisan conflict, this is no longer true today, as research has consistently shown that voters who identify as liberals are more liberal across multiple issue categories and that the same trend holds for conservatives [@abramowitz_great_2018]. As a result, whereas the prospect for cooperation on issues that were not politically polarized remained high in the past, today the prospect for cooperation in general is far lower [@layman_activists_2010].

A second challenge posed by rising polarization stems from the drastic increase of negative feelings toward those on the other side of the political spectrum, a phenomenon known as affective polarization or "negative partisanship"[@abramowitz_great_2018]. Just as contemporary polarization is marked by citizens who hold increasingly strong and negative feelings towards those of a different political stripe, there is likely a tendency for members of the military institution to harbor similar feelings and attitudes. One might object, pointing to the training and professional development military leaders receive that emphasizes the importance of upholding an apolitical and non-partisan ethic, but even a cursory review of the evidence (members of the military increasingly making presidential endorsements [@griffiths_not_2019] and posting political messages on social media [@urben_like_2017] ) suggests this is not the case.

A third, and I contend the most impactful, challenge posed by polarization to the normative conceptions of the military's role in politics stems from the fact that our polarized climate is marked by the belief among individuals that core issues - such as those that involve attitudes about race, religion, and multiple layers of not just individual, but also national identity - are at stake in the current polarized political climate. As the rise of "identity politics" illustrates, some of the most polarizing issues today are those that are related to an individual's identity, such as race, gender, and immigration [@sides_identity_2018]. And at the national level, the same phenomenon holds: as Vice President Joe Biden and President Donald Trump wind down the final days of their presidential campaigns, both claim that "the soul of the nation" is at stake [@dias_biden_2020].

## Military Prestige

With respect to the military institution itself, one of the most massive and influential changes that has occurred over the past roughly seventy years is the increase in overall prestige, credibility, and influence of the military institution. While key legislative acts passed since World War Two, to include the National Security Act of 1947 and the Goldwater-Nichols Act of 1946, may have fixed important gaps in America's ability to meet its national security requirements [@inboden_national_2017], they also cemented the military as one of if not the dominant institution and actor within American politics. Over the same time period, while there have been spikes as a result of exogenous events such as wars and scandals that have involved the military, public opinion towards the military has risen sharpy. Consider that in the late 1970s, as a likely result of public opinion over the Vietnam War, 54% of Americans indicated that they trusted the military "a great deal or quite a lot," whereas in 2019, 73% of Americas expressed the same sentiments [@gallup_polls_confidence_2007]. Unsurprisingly given these developments, compared to other government institutions, today the American public trusts the military to a remarkably high degree. According to a recent Pew Research Center poll, 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018].

While military leaders are rightfully cognizant of the benefits of holding the public's trust, a number of scholars and practitioners have noted that the trend of "military worship" is not only unhealthy [@fallows_tragedy_2014], but likely harmful for democracy [@friend_military_2018]. Esteeming the military more than it ought can perpetuate, according to Mara Karlin and Alice Friend, a number of harmful myths, including the idea that military service is somehow superior than other, non-military forms of service [@friend_military_2018]. Over the long haul, a military that is perceived to be more prestigious than its normative civilian superiors can, according to numerous scholars, be problematic for civilian control of the miltiary [@friend_military_2018].

Scholars have also shown that when elites possess a high degree of credibility, their ability to influence and shape public opinion is likewise high. In his influential study, _The Nature and Origins of Mass Opinion_, John Zaller argues that formations in mass public opinion are the result of three interacting components: "variation in the information carried in elite discourse, individual differences in attention to this information, and individual differences in political values and other predispositions" [@zaller_nature_1992]. Zaller argues that throughout modern America, most citizens, as non-elites, must rely heavily on the messages that come from elites regarding various topics and issue areas [@zaller_nature_1992]. Zaller further argues that the messages that elites send is necessarily "stereotyped" in some way: that is, the messages that elites send, in order to be understood by regular citizens who are part of the public, must be "short, simple, and highly thematic" [@zaller_nature_1992]. In developing this argument, Zaller is not claiming that regular citizens are uninformed or careless people who are incapable of processing complex information, but rather acknowledging that most citizens care most about a world that is not "beyond their personal experience" [@zaller_nature_1992]. Thus, his main point is simply that elites play a significant role in forming and swaying public opinion.

Therefore, as the public's trust in the military has grown in the years since Vietnam, so has the military's potential to impact public opinion. This is especially true of military elites, who hold substantial positions of authority. An increase in credibility of the institution may not indicate that the military or particular military elites will actually engage in behavior to deliberately impact public opinion, but rather that the potential to do so has increased, as the American people tend to listen to those whom they trust and revere.

# My Puzzle, Where it Fits, and Why It Matters

The central question I wish to explore in my dissertation is whether and why the US military, or portions of it, have become increasingly politically active, and whether such a shift portends a shift in the military's own "role beliefs" concerning how it relates to American government and society. As such, my dissertation will center on fundamental and longstanding debates about the concepts of authority, loyalty, and the proper role of the military with respect to government. What is different from previous works in the field is that I wish to explore changes in these concepts, particularly within the US military, over time and in historical perspective.

While there have been several vital studies, most of them incorporating original survey data, exploring the beliefs that military members have regarding forms of political behavior [@dempsey_our_2010; @urben_party_2013; @urben_wearing_2014], these studies have not sought to explain these changes. Similarly, there have also been a number of recent and important studies regarding the impacts of political polarization on the credibility society places in the military as an instition [@robinson_michael_danger_2018; @robinson_who_2019], as well as a host of impressionistic pieces noting the potentially detrimental impacts of polarization and politicization on the military institution [@feaver_we_2016; @barno_how_2016; @golby_jim_keep_2020; @reid_retired_2020 ]. However, these studies also have not empirically explored the impact of increasing polarization and institutioncal prestige on the military as an institution over time. I wish to close that gap, as I think doing so will help not only scholars, but also practitioners, to include members of the military, understand the answers to some very important questions.  These questions include:

* Among whom within the military institution is a shift in normative "role beliefs" occurring, and why?
* When did such a shift start, and why? 
* If there is a shift in the military's normative "role beliefs," how will that impact the military's organizational culture?
* Are some of the most polarized domestic issues compatible with historically normative conceptions of military professionalism, such as the principles of civilian supremacy and a non-partisan ethos? why or why not?
* What are the key challenges or obstacles to the military regaining and sustaining an effective non-partisan ethos?

Specifically, I desire to explore the relationship between polarizing domestic politics and institutional credibility on the professional identity and character of the military. In the post World War Two era in particular, the military's institutional power, social prestige, and level of credibility has generally grown, making the military a recognized and respected actor within American politics. And, especially since the 1970s, political polarization has also increased dramatically, to the point that America is now as polarized as it was at the end of the American Civil-War [@hare_polarization_2014]. I believe that these two trends, occurring at least in a broad sense simultaneously, have significantly impacted the military's beliefs about its proper role in politics, and that a systematic investigation of this relationship should be undertaken.

---
### Figure this out

@@

 In the past several years for example, there has been a concerted effort by the services to identify and punish senior leaders who engage in misconduct, to demonstrate to Congress that senior military leaders do not need assistance to handle sexual assault, and to otherwise prove that it can be trusted to accomplish its missions [@us_army_repairing_2014]. Former Chief of Staff of the Army, Retired General Ray Odierno, told fellow Army generals that "of great importance within the Army Profession is the duty of each of us to be a steward of our profession and to maintain its essential and hard won trust with society [Ray Odierno in @us_army_repairing_2014].

---

 I am specifically interested in the question of if and how political polarization impacts the concept of loyalty in the military. This is an interesting research question because it brings together at least three relevant strands of literatures which, when combined, generates an interesting set of puzzles and questions. First, one of the most fundamental normative hallmarks of the military's professional nature, at least by Huntingtonian standards, is that it ought to be an apolitical and non-partisan institution. Second, one of the features of the US military is that it prizes and elevates the trait of loyalty among those who serve and, by statute as well as normative tradition, exercises simultaneous dual loyalties to its civilian heads, the President as the Commander in Chief, and Congress. And third, we know from the American politics literature that ideological polarization and partisan loyalty are connected in that those citizens who hold to stronger ideological convictions tend to be the most loyal partisans in terms of voting and other political behaviors [@abramowitz_great_2018, 112-116].

@@
